Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees

### 2011-12-14 paper/draft1.tex

% UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)
Auto-tuning techniques are widely used and effective for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.
However, a drawback of this approach is that thorough or exhaustive
auto-tuning requires compiling many kernels and calling each one many times,
and this process is slow.  Furthermore, it is desirable to provide users
with unified library abstraction boundaries for operations such as image
filtering and matrix multiplication, even those these operations actually
correspond to a large set of potential problem configurations with a wide
variety of memory access patterns and computational bottlenecks.  How can we
draw on data from previous empirical auto-tuning of related problems on related
hardware to make a just-in-time implementation decision for a novel problem?
This paper presents a machine learning approach to auto-tuning, in which
features of the current hardware platform, the kernel configuration and the
problem instance are passed to a regression model (boosted regression trees)
which predicts how much faster this kernel will be than a reference
baseline.  Combinatorial optimization strategies for auto-tuning that would
normally require evaluating large number of kernel configurations on the real
hardware, are made orders of magnitude faster by evaluating the surrogate
regression model instead.  We validate our approach using the filterbank
correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
0.1 seconds of hill climbing on the regression model (``{\em
predictive} auto-tuning'') can achieve an average of 95\% of the
speed-up brought by minutes of empirical auto-tuning.  Our approach is not
specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
be applied to almost any templated-code optimization problem, spanning a wide
variety of problem types, kernel types, and platforms.
% UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)

>>>>

% UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)
The rapidly evolving landscape of multicore architectures makes
the construction of efficient libraries a daunting task.
A family of methods known collectively as ``auto-tuning'' has emerged to
address this challenge.  Two major approaches to auto-tuning are
{\em empirical} and {\em model-based}:
empirical auto-tuning is a generic but slow approach that works by measuring runtimes of
candidate implementations, model-based auto-tuning predicts those runtimes using simplified hardware abstractions designed by hand.
We show that machine learning methods for non-linear regression can be used
to estimate timing models from data, capturing the best of both approaches.
A statistically-derived model offers the speed of a model-based approach,
with the generality and simplicity of empirical auto-tuning.
We validate our approach using the filterbank
correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
0.1 seconds of hill climbing on the regression model (``{\em
predictive} auto-tuning'') can achieve an average of 95\% of the
speed-up brought by minutes of empirical auto-tuning.  Our approach is not
specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
be applied to almost any templated-code optimization problem, spanning a wide
variety of problem types, kernel types, and platforms.
% UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)
